The Astropy Project: \\ Sustaining and Growing a Community-oriented Open-source Project and the Latest Major Release (v5.0) of the Core Package

### 2022-03-01 src/ms.tex
Persistent storage section
Signed-off-by: nstarman <nstarkman@protonmail.com>

    Units were also integrated further within \astropypkg. For already-defined
    \astropyModel, for instance imported from another library, \astropymodeling
    can now add units to otherwise unit-less models.
    % TODO! transition words
    A new module – \astropycosmologyunits – has been added to the cosmology
    sub-package for defining and collecting cosmological units and
    equivalencies. The unit \texttt{littleh} and equivalency
    \astropyapidoc{cosmology.units.with_H0}{with\_H0} are deprecated from the
    main \astropyunits sub-package and moved to \astropycosmologyunits. A new
    unit, \texttt{redshift}, is added for tracking factors of cosmological
    redshift. To correctly use redshift units in dimensional analysis,
    \texttt{redshift} is
    \astropyapidoc{cosmology.units.dimensionless_redshift}{treated as
    dimensionless} by default. To convert between redshift and other
    cosmological distance measures, e.g. CMB temperature or comoving distance,
    the equivalency
    \astropyapidoc{cosmology.units.with_redshift}{with\_redshift}
    is also added. This equivalency is actually a composite of other
    equivalencies, which may be used separately.

@@

    % TODO: this is very rough. Probably needs a bit of a re-write.

@@

    \astropy believes however array-valued data is stored, it should be usable
    as a column in a table: historical differences are baked into stable API,
    computational needs can differ, and some objects have extra information. The
    \astropytable and \astropyfits table implementations allow for mixin-columns
    (in general a mixin column may contain multiple data components as well as
    object attributes beyond the standard column). A well defined protocol has
    been developed to allow the usage of these mixin columns while allowing the
    object to ``round-trip'' through tables with no loss of data or attributes.
    With this protocol, it is now possible to store not only standard
    \astropyTableColumn objects to a FITS table
    \href{https://docs.astropy.org/en/stable/io/fits/api/hdus.html}{HDU}, but
    also any \astropy native objects within a \astropyTable. This includes
    \astropyTime, \astropyQuantity, \astropySkyCoord, and many others. For
    objects not already covered by the mixin protocol, functions can be
    registered with \astropyTable to convert, on-the-fly, any array-like object
    into a `mixin' column that can be used in a table. As an example, this is
    used to provide the seamless integration of \package{dask} arrays in
    \astropyTable, allowing computer cluster-scale data to be stored in a
    \astropyTable.

@@

  \paragraph{Persistent Storage and I/O}

    % TODO! need good transition sentence. This one is bad.
    The column mixin options in tables are mirrored in the increased support for
    I/O options.

    \begin{itemize}
      \item ECSV : Support for multidimensional and object columns in ECSV
      \item FITS : Support for writing Dask arrays to FITS files, and Append table to existing FITS file
      \item MRT Support for reading and writing tables to MRT format : GSOC
      \item Support for reading and writing tables to Parquet format
      \item Support for reading and writing tables to QDP format
      \item ASDF read write
    \end{itemize}

    Also stuff that isn't a table

    \begin{itemize}
      \item Support for reading, writing, and converting Cosmology
    \end{itemize}

>>>>

    Units were also integrated further within \astropypkg. For an already-defined
    \astropyModel, for instance imported from another library, \astropymodeling
    can now add units to otherwise unit-less models.
    % TODO! transition words
    A new module – \astropycosmologyunits – has been added to the cosmology
    sub-package for defining and collecting cosmological units and
    equivalencies. The unit \texttt{littleh} and equivalency
    \astropyapidoc{cosmology.units.with_H0}{with\_H0} are deprecated from the
    main \astropyunits sub-package and moved to \astropycosmologyunits. A new
    unit, \texttt{redshift}, is added for tracking factors of cosmological
    redshift. To correctly use redshift units in dimensional analysis,
    \texttt{redshift} is
    \astropyapidoc{cosmology.units.dimensionless_redshift}{treated as
    dimensionless} by default. To convert between redshift and other
    cosmological distance measures, e.g. CMB temperature or comoving distance,
    the equivalency
    \astropyapidoc{cosmology.units.with_redshift}{with\_redshift}
    is also added. This equivalency is actually a composite of other
    equivalencies, which may be used separately.

@@

@@

    \astropy believes that however array-valued data is stored, it should still
    be usable as a column in a table. A well defined protocol for mixin-columns
    has been developed for \astropytable and \astropyfits table, allowing the
    original object to be used as a column with a familiar API, and to
    ``round-trip'' through tables with no loss of data or attributes. With this
    protocol, it is now possible to store standard \astropyTableColumn objects
    to a FITS table
    \href{https://docs.astropy.org/en/stable/io/fits/api/hdus.html}{HDU}, and
    also any \astropy native objects within a \astropyTable. This includes
    \astropyTime, \astropyQuantity, \astropySkyCoord, and many others. For
    objects not already covered by the mixin protocol, functions can be
    registered with \astropyTable to convert on-the-fly any array-like object
    into a mixin column. As an example, this is used to provide the seamless
    integration of \package{dask} arrays with \astropyTable, allowing cloud-stored
    or cluster-scale data to be used as a column in a \astropyTable.

@@

  \paragraph{Persistent Storage}

    \astropy's efforts to increase support for column types in a table is
    mirrored by efforts to expand storage format options. New formats have been
    added, and existing formats updated to support more column types. We detail
    the changes, starting with the new formats.

    \astropy now supports reading and writing tables in the American
    Astronomical Society Journals' Machine-Readable Table (MRT) format. This has
    been a long-missing ASCII format, and we are pleased that this year's Google
    Summer of Code project adds MRT to \astropyTable I/O.  
    For details see the
    \href{https://docs.astropy.org/en/stable/io/ascii/write.html#cds-mrt-format}{Machine-Readable
    Table Format} section. In addition, \astropyTable may also read and write to
    \href{https://asdf-standard.readthedocs.io/en/latest/}{ASDF},
    \href{https://docs.astropy.org/en/stable/io/unified.html#table-io-parquet}{Parquet},
    and \href{https://wwwastro.msfc.nasa.gov/qdp/}{QDP} formats.

    The \href{https://github.com/astropy/astropy-APEs/blob/main/APE6.rst}{ECSV
    standard} has been updated to version 1.0, adding support for reading and
    writing \astropyTable with three additional data subtypes. First,
    multidimensional column data (both masked and unmasked) with fixed
    dimensions in all table cells. Second, multidimensional column data with
    variable-dimension arrays similar to FITS variable-length arrays. Third,
    object-type columns with simple \python objects.

    As mentioned previously, \package{dask} arrays may be used as a mixin column
    in \astropyTable. \package{dask} may also be the data array
    \astropyapidoc{io.fits.PrimaryHDU}{FITS HDU} and if written to disk, the
    array will be computed while written, avoiding excessive memory use.
    \astropyTable can now be appended to an existing FITS file, and
    \package{dask} mixin columns interoperate seamlessly between the two table
    types.
